---
title: "Information theory provides a consistent framework for the analysis of spatial patterns"
date: "2019-05-23"
draft: yes
slug: ent-bp1
output:
  html_document:
    code_folding: show
tags:
- spatial
- geocompr
- sil
- rstats
- landscape-ecology
categories: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      fig.width = 10,
                      fig.height = 2.5)
```

<!--intro-->
<!-- information theory -->
<!-- landscape metrics -->
<!-- landscapemetrics -->

<!-- prep -->
<!--adds data download-->

To learn more about this topic, read our open access article:

> Nowosad, J., and T. F. Stepinski. "Information Theory as a consistent framework for quantification and classification of landscape patterns". Landscape Ecology, DOI: 10.1007/s10980-019-00830-x

```{r, message=FALSE, echo=FALSE}
library(raster)
all_examples_stack = stack("data/lc_examples_stack.grd")
forest_examples = all_examples_stack[[grep("forest", names(all_examples_stack))]]
forest_examples = subset(forest_examples, c(5, 16, 1, 13, 26, 35))
# forest_examples = forest_examples[[c(1, 13, 26, 35)]]
forest_examples[[1]][] = ifelse(forest_examples[[1]][] != 2, 1, forest_examples[[1]][])
forest_examples[[2]][] = ifelse(forest_examples[[2]][] != 2, 1, forest_examples[[2]][])
forest_examples[[5]][] = ifelse(forest_examples[[5]][] != 6, forest_examples[[5]][], 4)
forest_examples[[6]][] = ifelse(forest_examples[[6]][] != 2, forest_examples[[6]][] - 3, forest_examples[[6]][])
landscapes = brick(forest_examples)
```


```{r, eval=FALSE, echo=FALSE}
library(tmap)
tmap_mode("plot")
tm_shape(all_examples_stack[[grep("forest", names(all_examples_stack))]]) +
  tm_raster(palette = lc_palette, style = "cat") +
  tm_facets(free.scales = FALSE) +
  tm_layout(panel.labels = as.character(seq_len(raster::nlayers(forest_examples))))
```

# Example data

<!--adds data download-->
Let's consider six fairly simple, but diverse examples:

1. One dominating category and second spatially aggregated minor category
2. One dominating category and second spatially disaggregated minor category
3. One less dominating category and second spatially aggregated minor category
4. One less dominating category and second spatially disaggregated minor category
5. Four categories, where each one is spatially aggregated
6. Four categories, where each one is spatially disaggregated

```{r}
lc_palette = c("#FFFF64", "#006400", "#966400", "#BE9600")

library(tmap)
tm_shape(landscapes) +
  tm_raster(palette = lc_palette, style = "cat",
            title = "Land cover category", 
            labels = c("Argiculture", "Forest", "Shrubland", "Grassland")) +
  tm_facets(free.scales = FALSE, ncol = 6) +
  tm_layout(panel.labels = seq_len(raster::nlayers(landscapes)),
            legend.outside.position = "bottom")
```

The example data represent different land cover categories; however, they could be any discrete values.
Additionally, those cases are only a small subset of possible spatial arrangement, but the ideas presented below works for more and less complex cases.

# Spatial represetation

One of the ways to describe the above examples is to count a number of cells for each category (this metrics is known as a composition).
However, this representation does not contain any information about the spatial distribution of the values.
For example, let's consider example data 1 and 2 -- they both have a very similar number of cells for each category, but the spatial arrangement of the values is very different.

Another way to represent the above example is to calculate a co-occurrence matrix.^[Another name for this representation is the adjacency matrix.]
It is created by counting all of the pairs of the adjacent cells in the data.
As you will see below, this representation describes not only composition but also spatial configuration of the values.

[The **landscapemetrics** package](https://r-spatialecology.github.io/landscapemetrics/index.html) allows for calculating co-occurrence matrices using the `get_adjacencies()` function.

```{r}
library(landscapemetrics)
get_adjacencies(landscapes)
```

The result for each example dataset is a matrix with a number of rows and columns equal to the number of unique categories in the dataset.
For example data 1, 200 times cells of the first category are adjacent to other cells of this category, 3180 times cells of the second category are adjacent to other cells of this category, and 50 times cells of the first category are adjacent to cells of the second category.
This representation shows that the second category dominates in the data (3180 pairs), but also that the cells from the first category are more often next to other cells of this category than a different one (200 vs. 50).

# Information theory

<!-- ... -->

## Marginal entropy [H(x)]

The first metric, **marginal entropy**, represents a diversity (thematic complexity, configuration) of spatial categories.
It is calculated as the entropy of the marginal distribution.
The `lsm_l_ent` function allows for calculating **marginal entropy**.
Its output gives a layer number (example data id), informs that the data is only of the landscape level, this metric abbreviation is `ent`, and finally it shows the resulting value.

```{r}
mar_ent = lsm_l_ent(landscapes)
mar_ent
```

<!-- The values can be vizualized together with spatial data using the **tmap** package. -->

```{r, echo=FALSE}
tm_shape(landscapes) +
  tm_raster(palette = lc_palette, style = "cat", legend.show = FALSE) +
  tm_facets(free.scales = FALSE, ncol = 6) +
  tm_layout(panel.labels = round(mar_ent$value, 2), 
            panel.label.size = 2,
            main.title = "Marginal entropy [H(x)]")
```

The resulting values indicate that the example datasets have different levels of thematic complexity.
Example data 1 and 2 have one dominating category (low values of **marginal entropy**), while categories are more evenly distributed in example data 3 and 4 (medium values of **marginal entropy**).
Example data 5 and 6 have the highest levels of thematic complexity due to the fact of having more unique, evenly distributed categories.

## Conditional entropy [H(y|x)]

The second metric, **conditional entropy**, represents a configurational complexity (geometric intricacy) of a spatial pattern. 
If the value of **conditional entropy** is small cells of one category are predominantly adjacent to only one class of cells.
On the other hand, the high value of **conditional entropy** shows that cells of one category are adjacent to cells of many different categories.
**Conditional entropy** can be calculated using the `lsm_l_condent` function.

```{r}
cond_ent = lsm_l_condent(landscapes)
cond_ent
```

```{r, echo=FALSE}
tm_shape(landscapes) +
  tm_raster(palette = lc_palette, style = "cat", legend.show = FALSE) +
  tm_facets(free.scales = FALSE, ncol = 6) +
  tm_layout(panel.labels = round(cond_ent$value, 2), 
            panel.label.size = 2,
            main.title = "Conditional entropy [H(y|x)]")
```

<!-- interpretation -->
<!-- The **conditional entropy** values are the smallest for example data 1, where most of the "green" cells are adaject to other "green" ones, and the "yellow" cells are next to other "yellow" cells. -->
<!-- In example data 2 is the "yellow" cells are adaject to -->
<!-- ... -->

## Joint entropy

The third metric, **joint entropy**, is an overall spatio-thematic complexity metric. 
It represents the uncertainty in determining a category of the focus cell and the category of  the adjacent cell.
In other words, it measures diversity of values in a co-occurence matrix -- the smaller the diversity, the value of **joint entropy** is larger.
You can use the `lsm_l_joinent()` to calculate this metric.

```{r}
join_ent = lsm_l_joinent(landscapes)
join_ent
```

```{r, echo=FALSE}
tm_shape(landscapes) +
  tm_raster(palette = lc_palette, style = "cat", legend.show = FALSE) +
  tm_facets(free.scales = FALSE, ncol = 6) +
  tm_layout(panel.labels = round(join_ent$value, 2), 
            panel.label.size = 2,
            main.title = "Joint entropy [H(x, y)]")
```

<!-- interpretation -->
<!-- ... -->
However, **joint entropy** alone is not capable to fully distinguish situations with spatially aggregated values (example data 1, 3, and 5) from situations that are spatially disaggregated (example data 2, 4, 6).

## Mutual information

The fourth metric, **mutual information**, quantifies the information that one cell in a pair provides about the other cell in a pair.
It disambiguates landscape pattern types characterize by the same value of an overall complexity.
The `lsm_l_mutinf()` function calculates **mutual information**.

```{r}
mut_inf = lsm_l_mutinf(landscapes)
mut_inf
```

```{r, echo=FALSE}
tm_shape(landscapes) +
  tm_raster(palette = lc_palette, style = "cat", legend.show = FALSE) +
  tm_facets(free.scales = FALSE, ncol = 6) +
  tm_layout(panel.labels = round(mut_inf$value, 2), 
            panel.label.size = 2,
            main.title = "Mutual information [H(x)]")
```

Larger values idicate that the cells of the same category are more aggregated, while the smaller values are indication of disaggregation.
The resulting values can be useful to differentiate between examples of similar **marginal entropy**.

## Relative mutual information

Due to the spatial autocorrelation, the value of mutual information tends to grow with a diversity of the landscape (**marginal entropy**).
To adjust this tendency, it is possible to calculate **relative mutual information** by dividing the **mutual information** by the **marginal entropy**.
**Relative mutual information** always has a range between 0 and 1 and can be used to compare spatial data with different number and distribution of categories.

```{r}
rel_mut_inf = lsm_l_mutinf(landscapes)$value / lsm_l_ent(landscapes)$value
rel_mut_inf
```

```{r, echo=FALSE}
tm_shape(landscapes) +
  tm_raster(palette = lc_palette, style = "cat", legend.show = FALSE) +
  tm_facets(free.scales = FALSE, ncol = 6) +
  tm_layout(panel.labels = round(rel_mut_inf, 2), 
            panel.label.size = 2,
            main.title = "Relative mutual information [U]")
```

**Relative mutual information** for the example data orders them from the least aggregated (example data 2), through those with medium aggregation (example data 6, 4, 5), to the most aggregated (example data 1 and 2).

# Applications

<!-- order -->
<!-- classify -->

# Summary

<!-- landscapemetrics -->
<!-- + the whole org -->

<!-- new paper -->
<!-- summary of the paper -->
<!-- + the whole sil org -->